ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2024-12-12 paper/Manuscript-Data-Package.Rmd
Edited heading before pushing back to Bruno
Detalling how to use the package in the data paper

```{r load-data, include = FALSE}
# Load data
data("gss_episodes")
data("gss_impedances")
```

@@

For {ActiveCA}, data we extracted all data needed to characterize active travel 
in Canada, namely, episodes where the activity was identified as moving between 
an origin and a destination, either by walking or cycling. Although Statistics 
Canada offers Public Use Microdata Files and documentation for the GSS program 
[see @statisticscanada2024], accessing these files, and preparing  them for analysis
is not a straightforward matter, given their size and complexity.  The  process 
of extracting information of interest from the source files is  time-consuming, 
tedious, and challenging and/or prone to error due to the expertise required to 
work with these files. To create {ActiveCA} we collected, cleaned, and processed
the cycles of the GSS  surveys concerning time use to make them ready for analysis.

@@

<!-- 
is an `R` data package that provides analysis-ready data from the General Social Survey of Canada  [@statisticscanada2024] regarding active travel. These data can be used to study walking and cycling in Canada for various purposes over time, and to calculate impedance functions for accessibility analysis. The package provides pre-estimated impedance functions for cycling and walking trips for Census Agglomerations and Census Metropolitan Areas.
-->

@@

<!-- 
Maybe here provide an example of the structure of this table?
-->

@@

<!-- 
Maybe here provide an example of the structure of this table?
-->

# Methods

For each selected cycle of the GSS surveys, we reviewed the episode files to 
identify episodes of movement that involved walking or cycling. This allowed us 
to select the activities immediately before and after the movement episode. After
that, we labeled the code variables with their appropriate descriptions, identifying
each origin and destination, mode of travel, as well the province and urban 
classification of the episode.

Show a graphical example of this process: how were active trips identified? 
What was the purpose of finding the preceding and succeeding activities?

Not needed in this paper:

Additionally, for Census Agglomerations and Census Metropolitan Areas we pre-estimated impedance functions for cycling and walking trips. Impedance functions reveals important information about the travel behavior of the population, by describing the relationship between the population at an origin and their likelihood or ability to travel to specific destinations to access opportunities [@soukhov2024].

Impedance functions are commonly used in accessibility analysis to calculate the cost of travel between different locations [@hansen1959; @páez2012; @palacios2022]. However, these functions need to be calibrated in order to accurately represent travel behavior. One effective way of calibrating an impedance function is by using the trip length distribution (TLD) obtained from origin-destination data [@soukhov2024]. The TLD represents the probability that a certain proportion of trips will be made at a specific cost, such as travel time. In this data set, low travel times are associated with a higher proportion of trips, while high travel times are associated with a lower proportion of trips.

For each combination of year, destination, and transportation mode (walking or cycling), we fitted the most suitable impedance function based on empirical data from the GSS surveys. We used the `fitdistrplus` package [@mullerdutang2015] to estimate the functions, selecting the distribution with the lowest Akaike information criterion (AIC) among exponential, gamma, log-normal, normal, and uniform types.

# How to use {ActiveCA}

The paper does not really need to report "Results". There are too many potential use cases and we would rather let users find their own results depending on what they are interested in. Instead of results, a section on "How to use the package" is appropriate. There are a number of points to highlight here, for example, how to load the package, call the data sets, in particular the use of episode weights to make sure that any statistical analysis is valid.

>>>>

```{r load-data, include = FALSE}
# Load data
data("gss_episodes")
data("gss_impedances")
data("gss_main_2015")

gss_m_2015 <- read.csv(paste0(here(), "/data-raw/source-files/Time-use-2015/gss-m.csv"))

gss_e_2015 <- read.csv(paste0(here(), "/data-raw/source-files/Time-use-2015/gss-e.csv"))
```

@@

For {ActiveCA} and using GSS Public Use Microdata Files (PUMFs), we extracted all data needed to characterize active travel 
in Canada, namely, episodes where the activity was identified as moving between 
an origin and a destination, either by walking or cycling. Although Statistics 
Canada offers Public Use Microdata Files and documentation for the GSS program 
[see @statisticscanada2024], accessing these files, and preparing  them for analysis
is not a straightforward matter, given their size and complexity.  The  process 
of extracting information of interest from the source files is  time-consuming, 
tedious, and challenging and/or prone to error due to the expertise required to 
work with these files. To create {ActiveCA} we collected, cleaned, and processed
the cycles of the GSS  surveys concerning time use to make them ready for analysis.

@@

@@

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first six variables of the GSS PUMF 2015 main file (Cycle 29). Each row in the table correspond to a survey respondent, while the columns refer the following information: record identification (`PUMFID`), the person's weight (`WGHT_PER`), the month the survey data was collected (`SURVMNTH`), the respondent's age group (`AGEGR10`), the respondent's sex (`SEX`), and the respondent's marital status (`MARSTAT`).

The main file of the 2015 GSS surveys includes a total of 17,390 respondents, representing `r sum(gss_m_2015$WGHT_PER)` individuals and 848 variables.  For discrete variables, Statistics Canada has assigned specific codes to the possible values, with each code accompanied by a label. For instance, in the case of the variable `SURVMNTH`, a value of `1` means `January 2016`, 2 means `February 2016`, 3 corresponds to `March 2016`, and so on.

As shown in Table \ref{tab:main-2015-unprocessed}, the variables are not labeled. Adittionaly, the format of the tables (comma-separated values) does not allow for the specification of variable types (whether a variable is continuous or discrete), which can lead to mistakes analysts who have limited experience working with PUMFs.

```{r gss-main-file-2015}
main_2015_unprocessed <- gss_m_2015[1:10, 1:6]  %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the main file of the 2015 GSS.") %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) %>%
  footnote(general = c("Legend: `PUMFID`: record identification. `WGHT_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent."), threeparttable = T)

main_2015_unprocessed
```

@@

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an example of the main file structure, Table \ref{tab:ep-2015-unprocessed} presents the first seven episodes for the record identification number `10041` and some variables from the GSS PUMF 2015 episode file (Cycle 29). Each row in the table corresponds to an episode associated with the specified record identification (`PUMFID = 10041`), including the episode's weight (`WGHT_EPI`), the episode number (`EPINO`), the activity code (`TUI_01`), the episode's duration (`DURATION`), and the episode's location (`LOCATION`).

In total, the episode file of the 2015 GSS surveys contains 274,108 records, representing `r sum(gss_e_2015$WGHT_EPI)` episodes and 527 variables that detail the episodes. Similar to the main file, Statistics Canada has created codes for the discrete variables, where each value has a corresponding label.

In the case illustrated in Table \ref{tab:ep-2015-unprocessed}, this respondent began the diary description while sleeping at home (`TUI_01 = 1` and `LOCATION = 300`) for 210 minutes, followed by 40 minutes of personal hygiene (`TUI_01 = 2`). The respondent then spent 15 minutes on personal care activities, such as getting ready for school, supervising homework, reading, playing, reprimanding, or providing educational or emotional support, as indicated by `TUI_01 = 27`. Next, they recorded a travel episode, walking for 15 minutes (`TUI_01 = 7` and `LOCATION = 315`), where the origin and destination were both their home. Such trips, where the journey starts and finishes at home, are classified as recreational and leisure trips.

In sequence, the respondent spent 3 hours searching for a job (`TUI_01 = 9`), took a 15-minute lunch break (`TUI_01 = 6`), and then cleaned the house (`TUI_01 = 18`) for two hours. Table \ref{tab:ep-2015-unprocessed} displays only six variables out of the 527 available. As shown, since the dataset does not label the variable's values, decoding the values for each variable can be a time-consuming and a challenging task.

```{r gss-epi-file-2015}
main_2015_unprocessed <- gss_e_2015 |>
  filter(PUMFID == 10041 & EPINO <= 7) %>%
  dplyr::select(PUMFID, EPINO,WGHT_EPI,TUI_01, DURATION, LOCATION) %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number `10041`.") %>%
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) %>%
  footnote(general = c("Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

main_2015_unprocessed
```

# Data extraction

For each selected cycle of the GSS surveys, we reviewed the episode files to identify episodes of movement that involved walking or cycling. This allowed us to select the activities immediately before and after the movement episode. After that, we labeled the code variables with their appropriate descriptions, identifying each origin and destination, mode of travel, time spent in the active trip, as well the province and urban classification of the episode.

The active trips (walking and cycling) were identified by the their correspondent activity code, considering the differences between the survey cycles. Identifying the proceeding and succeeding activities of each activity episode helps to understand the motivation of the trip and also enables to obtain the origin and destination of the trips.

For each GSS main files, we selected some socioeconomic variables for helping profiling the individuals engaged in active trip episodes. We selected valuable socioeconomic variables, such as age group of the respondent, sex of the respondent, marital status, number of children, between many others. After that, and similar to as we did to episode files, we labeled the code variables with their appropriate descriptions.

# How to use `ActiveCA`

The code chunk below demonstrates how you can install the `ActiveCA` in R:

```{r instal-ActiveCA}
if (!require("ActiveCA", character.only = TRUE)) {
      remotes::install_github("dias-bruno/ActiveCA")
}
```

After installing `ActiveCA` R package, you can load the package according to the following code:

```{r load-ActiveCA, warning=FALSE}
library(ActiveCA)
```

By running `?ActiveCA` in `R`, you can access all the package documentation. The documentation includes the description of the package, information about the authors and maintainers, and visualize all data sets already processed included in the package.

Figure X, display the documentation for the walking episodes of the 2015 GSS (`walking_2015.rda`). The documentation provides a brief description of the data set, how to load the package, also including the description of the format of the file and the description of all variables (already labeled).

After restart your R session, you can access the data set by loading each of them. The chunk below displays how to load 2015 GSS episodes cycles.

```{r load-ActiveCA, warning=FALSE}
load(walking_2015)
```

### 2024-12-12 paper/Manuscript-Data-Package.tex
Detalling how to use the package in the data paper

For \{ActiveCA\}, data we extracted all data needed to characterize
active travel in Canada, namely, episodes where the activity was
identified as moving between an origin and a destination, either by
walking or cycling. Although Statistics Canada offers Public Use
Microdata Files and documentation for the GSS program
\citep[see][]{statisticscanada2024}, accessing these files, and
preparing them for analysis is not a straightforward matter, given their
size and complexity. The process of extracting information of interest
from the source files is time-consuming, tedious, and challenging and/or
prone to error due to the expertise required to work with these files.
To create \{ActiveCA\} we collected, cleaned, and processed the cycles
of the GSS surveys concerning time use to make them ready for analysis.

@@

@@

\section{Methods}\label{methods}

>>>>

For \{ActiveCA\} and using GSS Public Use Microdata Files (PUMFs), we
extracted all data needed to characterize active travel in Canada,
namely, episodes where the activity was identified as moving between an
origin and a destination, either by walking or cycling. Although
Statistics Canada offers Public Use Microdata Files and documentation
for the GSS program \citep[see][]{statisticscanada2024}, accessing these
files, and preparing them for analysis is not a straightforward matter,
given their size and complexity. The process of extracting information
of interest from the source files is time-consuming, tedious, and
challenging and/or prone to error due to the expertise required to work
with these files. To create \{ActiveCA\} we collected, cleaned, and
processed the cycles of the GSS surveys concerning time use to make them
ready for analysis.

@@

The table \ref{tab:main-2015-unprocessed} shows the first ten lines and
the first 6 variables of the GSS PUMF 2015 main file (Cycle 29). Each
line of the table refers to a respondent in a survey and the columns
refer respectively to the record identification (\texttt{PUMFID}), the
person's weight (\texttt{WGHT\_PER}), the month the survey data was
collected (\texttt{SURVMNTH}), the respondent's age group
(\texttt{AGEGR10}), the respondent's sex (\texttt{SEX}), and the
respondent's marital status (\texttt{MARSTAT}). In total, the main file
of the 2015 GSS surveys has 17,390 respondents corresponding to
\ensuremath{2.9766399\times 10^{7}} people and 848 variables. For the
discrete variables, Statistic Canada has created codes for the possible
values and each value has its own label. For example, for the variable
\texttt{SURVMNTH}, the value \texttt{1} means \texttt{January\ 2016},
while 2 means \texttt{February\ 2016}, 3 is \texttt{March\ 2016} and so
on. As you can see from the example in Table
\ref{tab:main-2015-unprocessed}, the variables are not labeled. In
addition, the format of the tables (comma separated) does not allow you
to specify the type of each variable (whether it is continuous or
discrete, for example), which can cause confusion for an analyst with
relatively little experience in handling PUMFs.

\begingroup\fontsize{8}{10}\selectfont

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `WGHT\_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-main-file-2015}\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the main file of the 2015 GSS.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & 7 & 5 & 1 & 5\\
10001 & 8516.6140 & 7 & 5 & 1 & 1\\
10002 & 371.7520 & 1 & 4 & 2 & 1\\
10003 & 1019.3135 & 3 & 6 & 2 & 5\\
10004 & 1916.0708 & 9 & 2 & 1 & 6\\
\addlinespace
10005 & 1952.2015 & 4 & 1 & 1 & 6\\
10006 & 5761.5528 & 8 & 1 & 1 & 6\\
10007 & 466.0426 & 6 & 5 & 2 & 3\\
10008 & 2479.2991 & 2 & 2 & 2 & 1\\
10009 & 1436.1641 & 8 & 6 & 1 & 3\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

Similarly to Table \ref{tab:main-2015-unprocessed} that displayed an
example of the main file structure, Table \ref{tab:ep-2015-unprocessed}
shows the first seven episodes for the record identification number
\texttt{10041} and some variables of the GSS PUMF 2015 episode file
(Cycle 29). Each line of the table refers to an episode of the
especified record identification (\texttt{PUMFID\ =\ 10041}), the
episode's weight (\texttt{WGHT\_EPI}), the episode number
(\texttt{EPINO}), the activity code (\texttt{TUI\_01}), the episode's
duration (\texttt{DURATION}) and the episode's location
(\texttt{LOCATION}). In total, the episode file of the 2015 GSS surveys
has 274,108 records corresponding to \ensuremath{4.6183762\times 10^{8}}
episodes and 527 variables that detail the episodes. As similar to the
main file, Statistic Canada has created codes for the possible values of
the discrete variables with each value has its own label. Considering
the case displayed in Table \ref{tab:ep-2015-unprocessed}, this person
started the diary description while sleeping at home
(\texttt{TUI\_01\ =\ 1} and \texttt{LOCATION\ =\ 300}) for 210 minutes,
then engaged in personal hygiene for 40 minutes
(\texttt{TUI\_01\ =\ 2}), then spent 15 minutes getting personal care
(which can be getting ready for school, supervising homework, reading,
playing, reprimanding, educational, or emotional help, sinalized by the
\texttt{TUI\_01\ =\ 27}). After this, this person had an activity travel
episode, walking for 15 minutes (\texttt{TUI\_01\ =\ 7} and
\texttt{LOCATION\ =\ 315}) in which the origin and destination was their
own house. Cases like this, where the trip starts and finishes at home
refers to recreational and leisure trips. After that, the person spent 3
hours looking for a job (\texttt{TUI\_01\ =\ 9}), had break for a lunch
of 15 minutes duration (\texttt{TUI\_01\ =\ 6}) and cleaned the house
(\texttt{TUI\_01\ =\ 18}) for two hours. Table
\ref{tab:ep-2015-unprocessed} shows only six variables from the 527
possible cases. As you can see, since that data set is not labeled,
figuring out the codes for every variable can be time consuming and
difficult.

\begingroup\fontsize{8}{10}\selectfont

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-epi-file-2015}\label{tab:main-2015-unprocessed}Visualization of the seven first episodes of the record number `10041`.}\\
\toprule
PUMFID & EPINO & WGHT\_EPI & TUI\_01 & DURATION & LOCATION\\
\midrule
10041 & 1 & 1353.818 & 1 & 210 & 300\\
10041 & 2 & 1353.818 & 2 & 40 & 300\\
10041 & 3 & 1353.818 & 27 & 15 & 300\\
10041 & 4 & 1353.818 & 7 & 15 & 315\\
10041 & 5 & 1353.818 & 9 & 180 & 300\\
\addlinespace
10041 & 6 & 1353.818 & 6 & 15 & 300\\
10041 & 7 & 1353.818 & 18 & 120 & 300\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

\section{Data extraction}\label{data-extraction}
